---
title: "v014 teen births"
author: "how"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# about

## Basic information

-   Description: Number of births per 1,000 female population ages 15-19.

-   What is the numerator? The numerator is the total number of births to mothers ages 15-19 in the seven-year time period.

-   What is the denominator? The denominator is the aggregate female population, ages 15-19, over the seven-year time period.

## Raw data access

Population data come from Census population estimates (https://www2.census.gov/programs-surveys/popest/datasets) from 2018 to 2024. July 1 population estimates are used for each year (e.g., for 2022, set YEAR = 4 to get 7/1/2022 population estimate). Data files can be downloaded using the links below.

2024​ Census Population Estimates\
file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2024/cc-est2024-alldata.pdf ​\
Data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2024/counties/asrh/cc-est2024-alldata.csv ​

2023 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2023/cc-est2023-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/asrh/cc-est2023-alldata.csv

2022 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2022/cc-est2022-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2022/counties/asrh/cc-est2022-all.csv

2021 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2021/cc-est2021-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/counties/asrh/cc-est2021-all.csv

2020 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2020/cc-est2020-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/counties/asrh/CC-EST2020-ALLDATA.csv

2019 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2019/cc-est2019-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/asrh/cc-est2019-alldata.csv

2018 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2018/cc-est2018-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/counties/asrh/cc-est2018-alldata.csv

NOT NEEDED FOR THE 2026 RELEASE:

2017 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2017/cc-est2017-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2010-2017/counties/asrh/cc-est2017-alldata.csv

2022, 2023, 2024 Connecticut (CT) counties population: used 2021 CT county population for legacy county FIPS codes. 2021 CT population values were used to calculate both state and national values as well.

We calculate v014 teen births using natality data requested from the National Center for Health Statistics. However, CDC WONDER can also be used to pull teen birth rates: https://wonder.cdc.gov/natality-current.html

If using CDC WONDER, 1) select the most recent 7 years, 2) select Age of Mother 15-19 years, and 3) group results by county

# import packages

```{r}
library(tidyverse)
library(haven)
library(here)
```

# set up directory

```{r}
# set up directory that holds data needed for calculations
# NCHS mortality data, not allowed to share in GitHub
nchs_mort_dir <- "E:/CHRR/mortality raw"

# NCHS natality data, not allowed to share in GitHub
nchs_nata_dir <- "E:/CHRR/natality raw/rawdata"


```

# standard county and state fips

-   The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.

```{r }
fips <-  bind_rows(
  read_sas(here("inputs/county_fips_with_ct_old.sas7bdat")),
  read_sas(here( "inputs/state_fips.sas7bdat"))) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

```

Load population data from census

```{r}
d24 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2020-2024/counties/asrh/cc-est2024-alldata.csv")

d23 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/asrh/cc-est2023-alldata.csv")

d22 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2020-2022/counties/asrh/cc-est2022-all.csv")

d21 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/counties/asrh/cc-est2021-all.csv")

d20 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/counties/asrh/CC-EST2020-ALLDATA.csv")

d19 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/asrh/cc-est2019-alldata.csv")

d18 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/counties/asrh/cc-est2018-alldata.csv")

#d17 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2017/counties/asrh/cc-est2017-alldata.csv")


```

combine years and subset population data to include females between ages 15 and 19

```{r}
library(tidyverse)
df = list(d18, d19, d20, d21, d22, d23, d24)
dfsub = lapply(df, function(x) x %>% filter(AGEGRP == 4, YEAR == max(YEAR, na.rm = TRUE)) %>% 
               select(TOT_FEMALE, STATE, COUNTY))

# fix connecticut for 2022, 2023, and 2024: set ct pop values for 2022, 2023, 2024 to be equal to ct vals in 2021 
ct21 = dfsub[[4]] %>% filter(STATE == 9) #get vals from 2021 

dfsub[[5]] = rbind(dfsub[[5]] %>% filter(STATE != 9), ct21) #replace for 2023 
dfsub[[6]] = rbind(dfsub[[6]] %>% filter(STATE != 9), ct21) #replace for 2023 
dfsub[[7]] = rbind(dfsub[[7]] %>% filter(STATE != 9), ct21) #replace for 2024 



#get state level value for CT from 2022 
# this is the current cause of differences w ganhua - he used the 2022 data to calc a state value for CT while i did not...
#ct22= d22 %>% filter(AGEGRP == 4, YEAR == max(YEAR, na.rm = TRUE), STATE ==9) %>%                select(TOT_FEMALE, STATE, COUNTY) %>% summarize(TOT_FEMALE = sum(TOT_FEMALE))

#ct23= d23 %>% filter(AGEGRP == 4, YEAR == max(YEAR, na.rm = TRUE), STATE ==9) %>%                select(TOT_FEMALE, STATE, COUNTY) %>% summarize(TOT_FEMALE = sum(TOT_FEMALE))


#fix some old fipscodes 
dfsub_fixed = lapply(dfsub, function(x) mutate(x, COUNTY = case_when(
      STATE ==46 & COUNTY ==113 ~102,
      STATE ==2 & COUNTY ==270 ~158,
      STATE ==51 & COUNTY ==515 ~019,
      TRUE ~ COUNTY
      )))

dftot = dfsub_fixed %>% reduce(full_join, by = c("STATE", "COUNTY")) 
dftot = data.frame(dftot)

# convert all years to numeric so they can be summed 
dftot[c(1, 4:9)] <- sapply(dftot %>% select(-c("STATE", "COUNTY")),as.numeric)



dftot$STATE = stringr::str_pad(dftot$STATE, width = 2, side = "left", pad = "0")
dftot$COUNTY = stringr::str_pad(dftot$COUNTY, width = 3, side = "left", pad = "0")


# get total pop of teen girls 
dftot$popsums = rowSums(dftot %>% select(-c("STATE", "COUNTY")), na.rm = TRUE)

dfn= dftot %>% select("STATE", "COUNTY", "popsums") 

```

Calculate state and national sums

```{r}
dst = dfn %>% group_by(STATE) %>% 
  summarise(popsums = sum(popsums, na.rm = TRUE))
dst$COUNTY = "000"

dntl = dfn %>% 
  summarise(popsums = sum(popsums, na.rm = TRUE))
dntl$COUNTY = "000" 
dntl$STATE = "00"


pop_all = rbind(dst, dntl, dfn)

#add fipscodes
pop_all_f = merge(fips, pop_all, by.y = c("STATE", "COUNTY"), by.x = c("statecode", "countycode"), all.x = TRUE)

```

# Load the NCHS natality data

```{r}


#nat17 = readr::read_fwf(paste0(nchs_nata_dir, "/NATL2017US.AllCnty.txt"),                          col_positions = readr::fwf_cols(county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat18 = readr::read_fwf(paste0(nchs_nata_dir, "/NATL2018US.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat19 = readr::read_fwf(paste0(nchs_nata_dir, "/nat2019us.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat20 = readr::read_fwf(paste0(nchs_nata_dir, "/NATL2020us.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat21 = readr::read_fwf(paste0(nchs_nata_dir, "/nat2021us.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat22 = readr::read_fwf(paste0(nchs_nata_dir, "/NATL2022US.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat23 = readr::read_fwf(paste0(nchs_nata_dir, "/nat2023us.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat24 = readr::read_fwf(paste0(nchs_nata_dir, "/nat2024us.AllCounties.txt"), 
                         col_positions = readr::fwf_cols(county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat = list( nat18, nat19, nat20, nat21, nat22, nat23, nat24)

#remove births with mother's residence American Samoa, Guam, Northern Marianas, Puerto Rico, Virgin Islands, not applicable, and not classified ; remove non teen births 
nat_fixed = lapply(nat, function(x) x %>% filter(mage == 2) %>% filter(!(state %in% c("AS", "GU", "MP", "PR", "VI", "XX", "ZZ", "AB", "BC", "MB", "NB", "ON", "QC", "SK", "ZZ")) & !(county_fips %in% c("000", "999"))) %>% mutate(birth = 1))


#fix some old fipscodes 
nat_fixed_again = lapply(nat_fixed, function(x) mutate(x, county_fips = case_when(
      state =="SD" & county_fips =="113" ~"102",
      state =="AK" & county_fips =="270" ~"158",
      state =="VA" & county_fips =="515" ~"019",
      TRUE ~ county_fips
      )))

natsums = lapply(nat_fixed_again, function(x) x %>% group_by(state, county_fips) %>% summarise(tbirth = sum(birth)))

nattot = natsums %>% reduce(full_join, by = c("state", "county_fips")) 
nattot = data.frame(nattot)
nattot[3:9] <- sapply(nattot[3:9],as.numeric)

nattot$birthsums = rowSums(nattot[3:9], na.rm = TRUE)



```

calculate state and national totals

this needs to be done before merging w fips to ensure that we don't inadvertently lose any births

```{r}

statesums = nattot %>% filter(county_fips != "000") %>% 
  group_by(state) %>% 
  summarise(birthsums = sum(birthsums, na.rm = TRUE))
statesums$county_fips = "000"

ntlsum = nattot %>% filter(county_fips != "000") %>% 
  summarise(birthsums = sum(birthsums, na.rm = TRUE))
ntlsum$state = "US" 
ntlsum$county_fips = "000"

# suppress counties with fewer than 10 teen births 
nattot$birthsums = ifelse(nattot$birthsums < 10, NA, nattot$birthsums)

nattot_county = nattot %>% filter(county_fips != "000") %>% 
  select(state, county_fips, birthsums)

natall = bind_rows(nattot_county, statesums, ntlsum)



```

add and format fipscodes select important cols only

```{r}

nattot_fips <- merge(fips, natall, by.x = c("state", "countycode"), by.y = c("state", "county_fips"), all.x = TRUE) 
```

Now create the final dataset and calculate rates

```{r}
#combine birth and population data 
births_pop = merge(nattot_fips, pop_all_f, by = c("statecode", "countycode"), all = TRUE)


births_pop$rate = births_pop$birthsums/ births_pop$pop*1000

#calculate poisson CIs
#(some of this code came from GL)
alo = (1-0.95)/2
ahi = (0.95 + 1)/2

dpoi = data.frame(n = 1:99)
for (i in 1:length(dpoi$n)) { 
dpoi$l = qgamma(alo, dpoi$n)/dpoi$n
dpoi$u = qgamma(ahi, dpoi$n+1)/dpoi$n
}


births_pop_ci = births_pop %>% 
  mutate(rse = 100*(sqrt(1/birthsums)), 
         stde = rate*rse/100,
         lci = ifelse(birthsums >= 100, rate - 1.96*stde, NA),
         uci = ifelse(birthsums >= 100, rate + 1.96*stde, NA))


births_pop_ci_p = merge(births_pop_ci, dpoi, by.x = "birthsums", by.y = "n", all.x = TRUE)

births_pop_ci_p$lci = ifelse(is.na(births_pop_ci_p$lci), births_pop_ci_p$l * births_pop_ci_p$rate, births_pop_ci_p$lci)
births_pop_ci_p$uci = ifelse(is.na(births_pop_ci_p$uci), births_pop_ci_p$u * births_pop_ci_p$rate, births_pop_ci_p$uci)

how = births_pop_ci_p


v014 = births_pop_ci_p %>% rename(v014_rawvalue = rate, 
                                  v014_cilow = lci, 
                                  v014_cihigh = uci,
                                  v014_numerator = birthsums, 
                                  v014_denominator = popsums) %>% 
  select(statecode, 
         countycode, 
         v014_numerator, 
         v014_denominator, 
         v014_rawvalue,
         v014_cihigh,
         v014_cilow)



```

Compare to GL

```{r}
compare("v014", v014, year = 2026)
```

save the measure_dataset

```{r}
#save to project 
write.csv(v014, file = here("measure_datasets/v014_R2026.csv"), row.names = FALSE)

```
