---
title: "v014 teen births - race/ethnicity subgroup"
author: "how"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# about

## Basic information

-   Description: Number of births per 1,000 female population ages 15-19.

-   What is the numerator? The numerator is the total number of births to mothers ages 15-19 in the seven-year time period.

-   What is the denominator? The denominator is the aggregate female population, ages 15-19, over the seven-year time period.

## Raw data access

Population data come from Census population estimates (https://www2.census.gov/programs-surveys/popest/datasets) from 2018 to 2024. July 1 population estimates are used for each year (e.g., for 2022, set YEAR = 4 to get 7/1/2022 population estimate). Data files can be downloaded using the links below.

2024​ Census Population Estimates\
file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2024/cc-est2024-alldata.pdf ​\
Data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2024/counties/asrh/cc-est2024-alldata.csv

2023 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2023/cc-est2023-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/asrh/cc-est2023-alldata.csv

2022 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2022/cc-est2022-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2022/counties/asrh/cc-est2022-all.csv

2021 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2021/cc-est2021-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/counties/asrh/cc-est2021-all.csv

2020 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2020/cc-est2020-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/counties/asrh/CC-EST2020-ALLDATA.csv

2019 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2019/cc-est2019-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/asrh/cc-est2019-alldata.csv

2018 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2018/cc-est2018-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/counties/asrh/cc-est2018-alldata.csv

NOT NEEDED FOR 2026 calculations: 2017 Census Population Estimates file layout: https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2010-2017/cc-est2017-alldata.pdf Data: https://www2.census.gov/programs-surveys/popest/datasets/2010-2017/counties/asrh/cc-est2017-alldata.csv

2022, 2023, 2024 Connecticut (CT) counties population: used 2021 CT county population for legacy county FIPS codes. 2021 CT population values were used to calculate both state and national values as well.

Subgroups are based on single-race categories.

We calculate v014 teen births using natality data requested from the National Center for Health Statistics. However, CDC WONDER can also be used to pull teen birth rates: https://wonder.cdc.gov/natality-current.html

If using CDC WONDER, 1) select the most recent 7 years, 2) select Age of Mother 15-19 years, and 3) group results by county

## Subgroups

Race/ethnicity is defined as follows:

if MRACEHISP = 1 then race=1; /\* Non-Hispanic White (only) */ if MRACEHISP = 2 then race=2; /* Non-Hispanic Black (only) */ if MRACEHISP = 3 then race=3; /* Non-Hispanic AIAN (only) */ if MRACEHISP = 4 then race=4; /* Non-Hispanic Asian (only) */ if MRACEHISP = 5 then race=5; /* Non-Hispanic NHOPI (only) */ if MRACEHISP = 6 then race=6; /* Non-Hispanic more than one race */ if MRACEHISP = 7 then race=8; /* Hispanic \*/

MRACEHISP: Mother’s Race/Hispanic Origin

# import packages

```{r}
library(tidyverse)
library(haven)
library(here)
```

# set up directory

```{r}
# set up directory that holds data needed for calculations
# NCHS mortality data, not allowed to share in GitHub
nchs_mort_dir <- "E:/CHRR/mortality raw"

# NCHS natality data, not allowed to share in GitHub
nchs_nata_dir <- "E:/CHRR/natality raw/rawdata"
```

# standard county and state fips

-   The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.

```{r }
fips <-  bind_rows(
  read_sas(here("inputs/county_fips_with_ct_old.sas7bdat")),
  read_sas(here( "inputs/state_fips.sas7bdat"))) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

```

Load population data from census

```{r}
d24 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2020-2024/counties/asrh/cc-est2024-alldata.csv")

d23 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/asrh/cc-est2023-alldata.csv")

d22 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2020-2022/counties/asrh/cc-est2022-all.csv")

d21 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/counties/asrh/cc-est2021-all.csv")

d20 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/counties/asrh/CC-EST2020-ALLDATA.csv")

d19 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/asrh/cc-est2019-alldata.csv")

d18 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2018/counties/asrh/cc-est2018-alldata.csv")

#d17 = read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2017/counties/asrh/cc-est2017-alldata.csv")


```

combine years and subset population data to include females between ages 15 and 19

```{r}
library(tidyverse)
df = list(d18, d19, d20, d21, d22, d23, d24)
dfsub = lapply(df, function(x) x %>% filter(AGEGRP == 4, YEAR == max(YEAR, na.rm = TRUE)) %>% 
                 select(NHWA_FEMALE, NHBA_FEMALE, NHIA_FEMALE, NHAA_FEMALE, NHNA_FEMALE, NHTOM_FEMALE, H_FEMALE, STATE, COUNTY))

# fix connecticut for 2022 and 2023: set ct pop values for 2022 and 2023 to be equal to ct vals in 2021 
ct21 = dfsub[[4]] %>% filter(STATE == 9) #get vals from 2021 
dfsub[[5]] = rbind(dfsub[[5]] %>% filter(STATE != 9), ct21) #replace for 2022 
dfsub[[6]] = rbind(dfsub[[6]] %>% filter(STATE != 9), ct21) #replace for 2023 
dfsub[[7]] = rbind(dfsub[[7]] %>% filter(STATE != 9), ct21) #replace for 2024 



#get state level value for CT from 2022 
#ct22= d22 %>% filter(AGEGRP == 4, YEAR == max(YEAR, na.rm = TRUE), STATE ==9) %>% select(TOT_FEMALE, STATE, COUNTY) %>% summarize(TOT_FEMALE = sum(TOT_FEMALE))

#ct23= d23 %>% filter(AGEGRP == 4, YEAR == max(YEAR, na.rm = TRUE), STATE ==9) %>% select(TOT_FEMALE, STATE, COUNTY) %>% summarize(TOT_FEMALE = sum(TOT_FEMALE))


#fix some old fipscodes 
dfsub_fixed = lapply(dfsub, function(x) mutate(x, COUNTY = case_when(
      STATE ==46 & COUNTY ==113 ~102,
      STATE ==2 & COUNTY ==270 ~158,
      STATE ==51 & COUNTY ==515 ~019,
      TRUE ~ COUNTY
      )))


dfsub_fixed_long = lapply(dfsub_fixed, function(x) x %>% pivot_longer(cols = NHWA_FEMALE: H_FEMALE, 
                                                  values_to = "pop",
                                                  names_to = "race"))


dftot = dfsub_fixed_long %>% reduce(full_join, by = c("STATE", "COUNTY", "race")) 
dftot = data.frame(dftot)

# convert all years of pop data to numeric so they can be summed 
dftot[4:10] <- sapply(dftot[4:10],as.numeric)


dftot$STATE = stringr::str_pad(dftot$STATE, width = 2, side = "left", pad = "0")
dftot$COUNTY = stringr::str_pad(dftot$COUNTY, width = 3, side = "left", pad = "0")


# get total pop of teen girls 
dftot$popsums = rowSums(dftot[4:10], na.rm = TRUE)

df_importantcols = dftot %>% select("STATE", "COUNTY", "race", "popsums") 

dfn = df_importantcols %>% group_by(STATE, COUNTY, race) %>% 
summarise(pop = sum(popsums, na.rm = TRUE))

```

Calculate state and national sums

```{r}
dst = dfn %>% group_by(STATE, race) %>% 
  summarise(pop = sum(pop, na.rm = TRUE))
dst$COUNTY = "000"

dntl = dfn %>% group_by(race) %>% 
  summarise(pop = sum(pop, na.rm = TRUE))
dntl$COUNTY = "000" 
dntl$STATE = "00"


pop_all = rbind(dst, dntl, dfn)


pop_all$racerecode = ifelse(pop_all$race == "NHWA_FEMALE", "white",
                            ifelse(pop_all$race == "NHBA_FEMALE", "black",
                                   ifelse(pop_all$race == "NHIA_FEMALE", "aian",
                                          ifelse(pop_all$race == "NHAA_FEMALE", "asian",
                                                 ifelse(pop_all$race == "H_FEMALE", "hispanic",
                                                        ifelse(pop_all$race == "NHNA_FEMALE", "nhopi", 
                                                        ifelse(pop_all$race == "NHTOM_FEMALE", "tom", NA)))))))




#add fipscodes
pop_all_f = merge(fips, pop_all, by.y = c("STATE", "COUNTY"), by.x = c("statecode", "countycode"), all.x = TRUE)

```

# Now load the NCHS births data

```{r}

#nat17 = readr::read_fwf(paste0(nchs_nata_dir, "/NATL2017US.AllCnty.txt"),                          col_positions = readr::fwf_cols(mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat18 = readr::read_fwf(paste0(nchs_nata_dir, "/NATL2018US.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat19 = readr::read_fwf(paste0(nchs_nata_dir, "/nat2019us.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat20 = readr::read_fwf(paste0(nchs_nata_dir, "/NATL2020us.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat21 = readr::read_fwf(paste0(nchs_nata_dir, "/nat2021us.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat22 = readr::read_fwf(paste0(nchs_nata_dir, "/NATL2022US.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat23 = readr::read_fwf(paste0(nchs_nata_dir, "/nat2023us.AllCnty.txt"), 
                         col_positions = readr::fwf_cols(mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat24 = readr::read_fwf(paste0(nchs_nata_dir, "/nat2024us.AllCounties.txt"), 
                         col_positions = readr::fwf_cols(mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat = list(nat18, nat19, nat20, nat21, nat22, nat23, nat24)

#remove births with mother's residence American Samoa, Guam, Northern Marianas, Puerto Rico, Virgin Islands, not applicable, and not classified ; remove non teen births 
nat_fixed = lapply(nat, function(x) x %>% filter(mage == 2) %>% filter(!(state %in% c("AS", "GU", "MP", "PR", "VI", "XX", "ZZ", "AB", "BC", "MB", "NB", "ON", "QC", "SK", "ZZ")) & !(county_fips %in% c("000", "999"))) %>% mutate(birth = 1))


#fix some old fipscodes 
nat_fixed_again = lapply(nat_fixed, function(x) mutate(x, county_fips = case_when(
      state =="SD" & county_fips =="113" ~"102",
      state == "AK" & county_fips == "270" ~"158",
      state =="VA" & county_fips == "515" ~"019",
      TRUE ~ county_fips
      )))




natsum = lapply(nat_fixed_again, function(x) x %>% group_by(state, county_fips, mracehisp) %>% summarise(tbirth = sum(birth)))

nattot = natsum %>% reduce(full_join, by = c("state", "county_fips", "mracehisp")) 
nattot = data.frame(nattot)
nattot[4:10] <- sapply(nattot[4:10],as.numeric)

nattot$birthsums = rowSums(nattot[4:10], na.rm = TRUE)


#remove births from virgin islands
nattot= nattot %>% filter(state != "VI") %>% select("state", "county_fips", "mracehisp", "birthsums") 



```

calculate state and national totals

```{r}
statesums = nattot %>% group_by(state, mracehisp) %>% 
  summarise(birthsums = sum(birthsums, na.rm = TRUE))
statesums$county_fips = "000"

ntlsum = nattot %>% group_by(mracehisp) %>% 
  summarise(birthsums = sum(birthsums, na.rm = TRUE))
ntlsum$state = "US" 
ntlsum$county_fips = "000"

natall = bind_rows(nattot, statesums, ntlsum)

# suppress counties with fewer than 10 teen births 
natall$birthsums = ifelse(natall$birthsums < 10, NA, natall$birthsums)

```

add and format fipscodes select important cols only

```{r}
nattot_fips <- merge(fips, natall, by.x = c("state", "countycode"), by.y = c("state", "county_fips"), all.x = TRUE) %>% select(statecode, countycode, birthsums, mracehisp)

```

recode the race variable to match the pop data

```{r}
nattot_fips$racerecode = ifelse(nattot_fips$mracehisp == 1, "white", 
                           ifelse(nattot_fips$mracehisp == 2, "black",
                           ifelse(nattot_fips$mracehisp == 3, "aian",
                            ifelse(nattot_fips$mracehisp == 4, "asian",
                            ifelse(nattot_fips$mracehisp == 5, "nhopi",
                            ifelse(nattot_fips$mracehisp == 6, "tom",
                            ifelse(nattot_fips$mracehisp == 7, "hispanic", "")))))))


```

Now create the final dataset and calculate rates

```{r}
#combine birth and population data 
births_pop = merge(nattot_fips, pop_all_f, by = c("statecode", "countycode", "racerecode"), all = TRUE)


births_pop$rate = births_pop$birthsums/ births_pop$pop*1000

#calculate poisson CIs
#(some of this code came from GL)
alo = (1-0.95)/2
ahi = (0.95 + 1)/2

dpoi = data.frame(n = 1:99)
for (i in 1:length(dpoi$n)) { 
dpoi$l = qgamma(alo, dpoi$n)/dpoi$n
dpoi$u = qgamma(ahi, dpoi$n+1)/dpoi$n
}


births_pop_ci = births_pop %>% 
  mutate(rse = 100*(sqrt(1/birthsums)), 
         stde = rate*rse/100,
         lci = ifelse(birthsums >= 100, rate - 1.96*stde, NA),
         uci = ifelse(birthsums >= 100, rate + 1.96*stde, NA))


births_pop_ci_p = merge(births_pop_ci, dpoi, by.x = "birthsums", by.y = "n", all.x = TRUE)

births_pop_ci_p$lci = ifelse(is.na(births_pop_ci_p$lci), births_pop_ci_p$l * births_pop_ci_p$rate, births_pop_ci_p$lci)
births_pop_ci_p$uci = ifelse(is.na(births_pop_ci_p$uci), births_pop_ci_p$u * births_pop_ci_p$rate, births_pop_ci_p$uci)


births_pop_wide = births_pop_ci_p %>% select(statecode, countycode, racerecode, birthsums, pop, rate, lci, uci) %>% 
  pivot_wider(id_cols = c("statecode", "countycode"), names_from = racerecode, values_from = c(birthsums, pop, rate, lci, uci))

how = data.frame(births_pop_wide)
how <- tidyr::unnest(how, cols = everything())

# Rename columns to match naming convention
names(how) <- gsub("^rate_(.*)$", "v014_race_\\1", names(how))
names(how) <- gsub("^lci_(.*)$", "v014_race_\\1_cilow", names(how))
names(how) <- gsub("^uci_(.*)$", "v014_race_\\1_cihigh", names(how))

# select important cols only 
v014 = how %>% select(statecode, countycode, 
                      v014_race_white, 
                      v014_race_white_cilow,
                      v014_race_white_cihigh, 
                      v014_race_black,
                      v014_race_black_cilow,
                      v014_race_black_cihigh, 
                      v014_race_asian,
                      v014_race_asian_cilow,
                      v014_race_asian_cihigh,
                      v014_race_aian,
                      v014_race_aian_cilow,
                      v014_race_aian_cihigh, 
                      v014_race_nhopi,
                      v014_race_nhopi_cilow,
                      v014_race_nhopi_cihigh,
                      v014_race_tom,
                      v014_race_tom_cilow,
                      v014_race_tom_cihigh, 
                      v014_race_hispanic, 
                      v014_race_hispanic_cilow,
                      v014_race_hispanic_cihigh)
```

compare to gl

```{r}
compare_subgroup("v014", v014, year = 2026)
```

save the measure_dataset

```{r}
#save to project 
write.csv(v014, file = here("measure_datasets/v014_otherdata_R2026.csv"), row.names = FALSE)

```
