---
title: "v024 children in poverty subgroups"
author: how 
format: html
---
# Description 
Percentage of people under age 18 in poverty.

# Connection to health 
Poverty and other social factors contribute to a number of deaths comparable to leading causes of death in the United States like heart attacks, strokes, and lung cancer.1 While repercussions resulting from poverty are present at all ages, children in poverty may experience lasting effects on academic achievement, health, and income into adulthood. Children living in households with low incomes have an increased risk of injury as a result of unsafe environments and are susceptible to more frequent and severe chronic conditions and their complications, such as asthma, obesity, diabetes, ADHD, behavior disorders, and anxiety, than children living in households with higher incomes.2-4

# Children In Poverty is a percentage 
Children in Poverty is the percentage of people under age 18 living in poverty.

# Children In Poverty is created using statistical modeling 
Data come from the Small Area Income and Poverty Estimates program, which uses data from the American Community Survey; estimates are produced using complex statistical modeling. Using modeling allows the generation of more stable estimates for places with small population or survey counts.

# Children In Poverty by race and ethnicity uses a different data source than overall county estimates 
We report Children in Poverty using the ACS race data categories: American Indian & Alaska Native, Asian & Pacific Islander, Black, Hispanic, and non-Hispanic White. The race and ethnicity data come from the American Community Survey using combined five-year estimates.

# Caution should be used when comparing these estimates across states 
Caution should be used when comparing SAIPE model-based estimates for different states or different counties in the same year. For more information please visit General Cautions About Comparing Estimates.

# Caution should be used when comparing these estimates across years 
Caution should be used when comparing SAIPE model-based estimates for different years. For more information please visit General Cautions About Comparing Estimates.

# Numerator 
The numerator is the number of people under age 18 living in a household whose income is below the poverty level. Poverty status is defined by family; either everyone in the family is in poverty or no one in the family is in poverty. The characteristics of the family used to determine the poverty threshold are: number of people, number of related children under 18, and whether or not the primary householder is over age 65. Family income is then compared to the poverty threshold; if that family’s income is below that threshold, the family is in poverty.

# Denominator 
The denominator is the total number of people under age 18 in a county.

# Data System
Small Area Income and Poverty Estimates; American Community Survey, five-year estimates

# Data Source
Administrative records, intercensal population estimates, decennial census, and American Community Survey (single-year, county-level observations); American Community Survey (five-year estimates)


# What years of data does this measure actually represent? 
2023 & 2019-2023

# What release year(s) of the data system were used? 
2024

# Any age or gender specifications?
Children under the age of 18

# If error margins are available for this measure, how are they calculated?
From SAIPE: "One goal of our small area estimation work is to provide estimates of the uncertainty surrounding the estimates of the numbers of people in poverty. The model-based estimates shown in the tables are accompanied by their 90-percent confidence intervals. These intervals were constructed from estimated standard errors. For the model-based estimates, the standard error depends mainly on the uncertainty about the model and the ACS sampling variance. While the variance of the shrinkage weights could also be a significant component of uncertainty about our estimates (if sizeable and ignored, we would be underestimating the standard errors), our research indicates that its contribution is negligible. (https://www.census.gov/programs-surveys/saipe/technical-documentation/methodology/counties-states/county-level.html)" 

# Is the measure age-adjusted?
No

# How are the units assigned to counties for this measure?
Residents who are living at the sample address at the time of the interview and who have lived or plan to live at that housing unit for more than 2 months are in scope for the survey

# How do we calculate this measure?
Create 95% confidence intervals from the given 90% confidence intervals.

cidistance_pov = 1.96/1.645*(POV_90UCL-POV_U18PCT);
v024_cilow =  v024_rawvalue - cidistance_pov;
v024_cihigh =  v024_rawvalue + cidistance_pov;

# Website to downloaded data 
https://www.census.gov/data/datasets/2023/demo/saipe/2023-state-and-county.html
Directions Orphaned in CHR&R Database
Create 95% confidence intervals from the given 90% confidence intervals.

# 95% Confidence Interval: 
see this document for formula: https://www2.census.gov/programs-surveys/acs/tech_docs/statistical_testing/2023_Instructions_for_Stat_Testing_ACS.pdf

cidistance_pov = 1.96/1.645*(POV_90UCL-POV_U18PCT);
v024_cilow =  v024_rawvalue - cidistance_pov;
v024_cihigh =  v024_rawvalue + cidistance_pov

# Online documentation 
https://www.census.gov/programs-surveys/saipe/technical-documentation/methodology/counties-states/county-level.html

# Can this measure progress? 
Modeled estimates have specific drawbacks with their usefulness in tracking progress in communities. Modeled data may not capture the effects of local conditions, such as health promotion policies. In order to better understand and validate modeled estimates, it can be helpful to supplement with additional local data. Additionally, it is important to examine national trends as you are assessing change in your own community.

# Finding More Data 
We recommend starting with the Small Area Income & Poverty Estimates website, which contains information on poverty by age and gender. Another resource is the Community Commons Health Equity Assessment Report, which maps children in poverty at the census tract level. You will need to log in to access the Health Equity Assessment Report, but the registration process is simple.

You can calculate poverty status by age and race using tables B17020A-G. For many communities, you can access the same tables at the census tract or census block level.

# Are subgroup estimates calculated?
Yes

# How are the subgroup data for this measure calculated?
Use ACS 5-year tables: B17020C (AI/AN), B17020D and B17020E (Asian and other Pacific Islander), B17020B (Black), B17020I (Hispanic), and B17020H (White, not Hispanic)

If a subgroup rate is 0% or 100%, the rate and the CI is suppressed. If a subgroup rate CI (prior to truncation) does not include any plausible value (e.g. cilow < 0% AND cihigh >100%), then the raw value and CI are also suppressed. This was done, as there are instances where rates are based on very small values, and CI ranges prior to truncation reflect a large magnitude of uncertainty.


# import standard packages

```{r}
library(tidyverse)
library(haven)
library(here)
library(tidycensus)
```


### standard county and state fips

The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.

```{r}
fips <-  bind_rows(
  read_sas(here("inputs/county_fips_with_ct_old.sas7bdat")) ,
  read_sas(here("inputs/state_fips.sas7bdat")))  %>% 
  arrange(statecode, countycode) %>% 
  glimpse()
```

# Load raw data 
We've chosen to use the tidycensus package to pull data from ACS directly into R. Alternatively, data can be downloaded from the ACS website at the link provided above 
```{r}
tables <- c(
  "B17020B", # Black
  "B17020C", # AIAN
  "B17020D", # Asian
  "B17020E", # NHPI
  "B17020H", # White NH
  "B17020I"  # Hispanic
)

county_data <- map_dfr(
  tables,
  ~ get_acs(
      geography = "county",
      table = .x,
      year = 2023,
      survey = "acs5"
    )
)

state_data <- map_dfr(
  tables,
  ~ get_acs(
      geography = "state",
      table = .x,
      year = 2022,
      survey = "acs5"
    )
) |> mutate(GEOID = str_pad(GEOID, width = 5, side = "right", pad = "0"))


ntl_data <- map_dfr(
  tables,
  ~ get_acs(
      geography = "us",
      table = .x,
      year = 2023,
      survey = "acs5"
    )) |> 
        mutate(GEOID = "00000")

poverty_race <- bind_rows(county_data, state_data, ntl_data)


# add labels and fipscodes, keep wide format  
poverty_race <- poverty_race %>%
  pivot_wider(
    id_cols = c(GEOID, NAME),
    names_from = variable,
    values_from = c(estimate, moe),
    names_glue = "{variable}{ifelse(.value == 'estimate','E','M')}"
  ) |> 
    mutate(fipscode = str_pad(GEOID, 5, side = "left", pad = "0"),
    statecode = substr(fipscode, 1, 2),
    countycode = substr(fipscode, 3, 5)
  )

```

# calculate numerators and denominators 

```{r}
# Calculate variables
pov_vars <- poverty_race %>%
  mutate(
    aiandenom = B17020C_011E + B17020C_012E + B17020C_013E + B17020C_003E + B17020C_004E + B17020C_005E, 
    aiannum = B17020C_003E + B17020C_004E + B17020C_005E, 
    v024aian = ifelse(aiannum/aiandenom != 0 & aiannum/aiandenom !=1, aiannum / aiandenom, NA), 
    aiansedenom = sqrt(B17020C_011M^2 + B17020C_012M^2 + B17020C_013M^2 + B17020C_003M^2 + B17020C_004M^2 + B17020C_005M^2) / 1.645,
    aiansenum = sqrt(B17020C_003M^2 + B17020C_004M^2 + B17020C_005M^2) / 1.645,
    aianse = ifelse((aiansenum^2 - (v024aian^2) * (aiansedenom^2)) > 0,
                    (1 / aiandenom) * sqrt(aiansenum^2 - (v024aian^2) * (aiansedenom^2)),
                    (1 / aiandenom) * sqrt(aiansenum^2 + (v024aian^2) * (aiansedenom^2))
    ),
    v024aianlow = ifelse(v024aian- 1.96* aianse <0 & v024aian+ 1.96* aianse>1, NA, pmax(v024aian - 1.96 * aianse, 0)),
    v024aianhigh = ifelse(v024aian- 1.96* aianse <0 & v024aian+ 1.96* aianse>1, NA, pmin(v024aian + 1.96 * aianse, 1)),
    v024aian = ifelse(v024aian- 1.96* aianse <0 & v024aian+ 1.96* aianse>1, NA, v024aian), 
    
    hispanicdenom = B17020I_011E + B17020I_012E + B17020I_013E + B17020I_003E + B17020I_004E + B17020I_005E,
    hispanicnum = B17020I_003E + B17020I_004E + B17020I_005E,
    v024hispanic = ifelse(hispanicnum/hispanicdenom != 0 & hispanicnum/hispanicdenom !=1, hispanicnum / hispanicdenom, NA), 
    
    hispanicsedenom = sqrt(B17020I_011M^2 + B17020I_012M^2 + B17020I_013M^2 + B17020I_003M^2 + B17020I_004M^2 + B17020I_005M^2) / 1.645,
    hispanicsenum = sqrt(B17020I_003M^2 + B17020I_004M^2 + B17020I_005M^2) / 1.645,
    hispanicse = ifelse((hispanicsenum^2 - (v024hispanic^2) * (hispanicsedenom^2)) > 0,
                        (1 / hispanicdenom) * sqrt(hispanicsenum^2 - (v024hispanic^2) * (hispanicsedenom^2)),
                        (1 / hispanicdenom) * sqrt(hispanicsenum^2 + (v024hispanic^2) * (hispanicsedenom^2))
    ),
    v024hispaniclow = ifelse(v024hispanic- 1.96* hispanicse <0 & v024hispanic+ 1.96* hispanicse>1, NA, pmax(v024hispanic - 1.96 * hispanicse, 0)),
    v024hispanichigh = ifelse(v024hispanic- 1.96* hispanicse <0 & v024hispanic+ 1.96* hispanicse>1, NA, pmin(v024hispanic + 1.96 * hispanicse, 1)),
    v024hispanic = ifelse(v024hispanic- 1.96* hispanicse <0 & v024hispanic+ 1.96* hispanicse>1, NA, v024hispanic), 
    
    
    
    whitedenom = B17020H_011E + B17020H_012E + B17020H_013E + B17020H_003E + B17020H_004E + B17020H_005E,
    whitenum = B17020H_003E + B17020H_004E + B17020H_005E,
    v024white = ifelse(whitenum/whitedenom != 0 & whitenum/whitedenom !=1, whitenum / whitedenom, NA), 
    
    whitesedenom = sqrt(B17020H_011M^2 + B17020H_012M^2 + B17020H_013M^2 + B17020H_003M^2 + B17020H_004M^2 + B17020H_005M^2) / 1.645,
    whitesenum = sqrt(B17020H_003M^2 + B17020H_004M^2 + B17020H_005M^2) / 1.645,
    whitese = ifelse((whitesenum^2 - (v024white^2) * (whitesedenom^2)) > 0,
                     (1 / whitedenom) * sqrt(whitesenum^2 - (v024white^2) * (whitesedenom^2)),
                     (1 / whitedenom) * sqrt(whitesenum^2 + (v024white^2) * (whitesedenom^2))
    ),
    v024whitelow = ifelse(v024white- 1.96* whitese <0 & v024white+ 1.96* whitese>1, NA, pmax(v024white - 1.96 * whitese, 0)),
    v024whitehigh = ifelse(v024white- 1.96* whitese <0 & v024white+ 1.96* whitese>1, NA, pmin(v024white + 1.96 * whitese, 1)),
    v024white = ifelse(v024white- 1.96* whitese <0 & v024white+ 1.96* whitese>1, NA, v024white), 
    
    
    blackdenom = B17020B_011E + B17020B_012E + B17020B_013E + B17020B_003E + B17020B_004E + B17020B_005E,
    blacknum = B17020B_003E + B17020B_004E + B17020B_005E,
    v024black = ifelse(blacknum/blackdenom != 0 & blacknum/blackdenom !=1, blacknum / blackdenom, NA), 
    
    blacksedenom = sqrt(B17020B_011M^2 + B17020B_012M^2 + B17020B_013M^2 + B17020B_003M^2 + B17020B_004M^2 + B17020B_005M^2) / 1.645,
    blacksenum = sqrt(B17020B_003M^2 + B17020B_004M^2 + B17020B_005M^2) / 1.645,
    blackse = ifelse((blacksenum^2 - (v024black^2) * (blacksedenom^2)) > 0,
                     (1 / blackdenom) * sqrt(blacksenum^2 - (v024black^2) * (blacksedenom^2)),
                     (1 / blackdenom) * sqrt(blacksenum^2 + (v024black^2) * (blacksedenom^2))
    ),
    v024blacklow = ifelse(v024black- 1.96* blackse <0 & v024black+ 1.96* blackse>1, NA, pmax(v024black - 1.96 * blackse, 0)),
    v024blackhigh = ifelse(v024black- 1.96* blackse <0 & v024black+ 1.96* blackse>1, NA, pmin(v024black + 1.96 * blackse, 1)),
    v024black = ifelse(v024black- 1.96* blackse <0 & v024black+ 1.96* blackse>1, NA, v024black), 
    asiandenom = B17020D_011E + B17020D_012E + B17020D_013E + B17020D_003E + B17020D_004E + B17020D_005E +
      B17020E_011E + B17020E_012E + B17020E_013E + B17020E_003E + B17020E_004E + B17020E_005E,
    asiannum = B17020D_003E + B17020D_004E + B17020D_005E + B17020E_003E + B17020E_004E + B17020E_005E,
    v024asian = ifelse(asiannum/asiandenom != 0 & asiannum/asiandenom !=1, asiannum / asiandenom, NA), 
    
    asiansedenom = sqrt(B17020D_011M^2 + B17020D_012M^2 + B17020D_013M^2 + B17020D_003M^2 + B17020D_004M^2 + B17020D_005M^2 +
                          B17020E_011M^2 + B17020E_012M^2 + B17020E_013M^2 + B17020E_003M^2 + B17020E_004M^2 + B17020E_005M^2) / 1.645,
    asiansenum = sqrt(B17020D_003M^2 + B17020D_004M^2 + B17020D_005M^2 + B17020E_003M^2 + B17020E_004M^2 + B17020E_005M^2) / 1.645,
    asianse = ifelse((asiansenum^2 - (v024asian^2) * (asiansedenom^2)) > 0,
                     (1 / asiandenom) * sqrt(asiansenum^2 - (v024asian^2) * (asiansedenom^2)),
                     (1 / asiandenom) * sqrt(asiansenum^2 + (v024asian^2) * (asiansedenom^2))
    ),
    v024asianlow = ifelse(v024asian- 1.96* asianse <0 & v024asian+ 1.96* asianse>1, NA, pmax(v024asian - 1.96 * asianse, 0)),
    v024asianhigh = ifelse(v024asian- 1.96* asianse <0 & v024asian+ 1.96* asianse>1, NA, pmin(v024asian + 1.96 * asianse, 1)),
    v024asian = ifelse(v024asian- 1.96* asianse <0 & v024asian+ 1.96* asianse>1, NA, v024asian), 
    
    
  ) %>%
  # Filter and clean state and county codes
  filter(statecode != "72") %>%
  mutate(
    statecode = ifelse(is.na(statecode), "00", statecode),
    countycode = ifelse(is.na(countycode), "000", countycode)
  ) %>% # select relevant cols 
  select(
    statecode,countycode, v024aian, v024aianlow, v024aianhigh, aiannum, aiandenom,
    v024hispanic, v024hispaniclow, v024hispanichigh, hispanicnum, hispanicdenom,
    v024white, v024whitelow, v024whitehigh, whitenum, whitedenom,
    v024black, v024blacklow, v024blackhigh, blacknum, blackdenom,
    v024asian, v024asianlow, v024asianhigh, asiannum, asiandenom
  )
```



# add in standardized fipscodes and rename  
```{r}
v024_subgroup = merge(fips, pov_vars, by = c("statecode", "countycode"), all.x = TRUE) 
```


# suppress a couple of counties with values outside expected/plausible
this is specific to the 2025 release. the decision to suppress the values for these subgroups and these counties was based on information gleaned during the validation step (not included in this measure caluclations repository)
```{r}
# some suppressions 
v024_subgroup <- v024_subgroup %>%
  mutate(
    # Set the specified Black variables to NA for specific statecode and countycode combinations
    v024black = ifelse((statecode == 51 & countycode == 580) | (statecode == 53 & countycode == "027"), NA, v024black),
    blacknum = ifelse((statecode == 51 & countycode == 580) | (statecode == 53 & countycode == "027"), NA, blacknum),
    blackdenom = ifelse((statecode == 51 & countycode == 580) | (statecode == 53 & countycode == "027"), NA, blackdenom),
    v024blacklow = ifelse((statecode == 51 & countycode == 580) | (statecode == 53 & countycode == "027"), NA, v024blacklow),
    v024blackhigh = ifelse((statecode == 51 & countycode == 580) | (statecode == 53 & countycode == "027"), NA, v024blackhigh),
    # Set the specified Hispanic variables to NA for a specific statecode and countycode combination
    v024hispanic = ifelse(statecode == 13 & countycode == 235, NA, v024hispanic),
    hispanicnum = ifelse(statecode == 13 & countycode == 235, NA, hispanicnum),
    hispanicdenom = ifelse(statecode == 13 & countycode == 235, NA, hispanicdenom),
    v024hispaniclow = ifelse(statecode == 13 & countycode == 235, NA, v024hispaniclow),
    v024hispanichigh = ifelse(statecode == 13 & countycode == 235, NA, v024hispanichigh)
  )

```

# save to project 
```{r}
readr::write_csv(v024_subgroup, file = here("measure_datasets/v024_subgroup_r2025.csv"))

```