---
title: "v037 low birth weight"
author: "Hannah Olson-Williams"
output: html_document
---

# Description 
Percentage of live births with low birth weight (< 2,500 grams).


# Connection to health 
Low birth weight can be due to preterm births (<37 weeks of gestation) and intrauterine growth restrictions that are associated with increased infant morbidity and mortality risks.1 Structural barriers to proper nutrition and adequate prenatal care can result in slowed intrauterine growth.2,3 Maternal stress and living in a neighborhood with a high level of pollution are associated with lower birth weight4 as well as substance misuse during pregnancy.5 Other factors that increase the likelihood of preterm births include multiple births, pre-eclampsia, and infections such as chorioamnionitis, bacterial vaginosis, and sepsis.2,3

Low birth weight is an important public health indicator that can be used to assess maternal health, nutrition, healthcare delivery, and poverty.1-3 Infants born with low birth weight have approximately 20 times greater chance of dying than those with normal birth weight.3 Infants who survive may face adverse health outcomes such as decreased growth, lower IQ, impaired language development, and chronic conditions (e.g., obesity, diabetes, cardiovascular disease) during adulthood.3,6

# Low Birth Weight is a percentage 
Low Birth Weight is the percentage of live births where the infant weighed less than 2,500 grams (approximately 5 lbs., 8 oz.).

# Births are counted in the mother's county of residence 
Births are counted in the county corresponding to the mother’s address on the child’s birth certificate, regardless of where the child was born.

# Some data are suppressed 
A missing value is reported for counties with fewer than 10 low birth weight births in the time frame.

# Caution should be used when comparing these estimates across years
Updates to race categories in the 2024 Annual Data Release mean that comparisons with previous years should be made with caution.

# Numerator 
The numerator is the number of live births for which the infant weighed less than 2,500 grams (approximately 5 lbs., 8 oz.) over seven years.

# Denominator 
The denominator is the total number of live births for which weight was recorded over seven years.

# Website to downloaded data 
https://wonder.cdc.gov/natality-current.html

# Directions
(The following steps are good to check internal calculations. But, actual calculation and duplication should use requested NCHS data.)

Select the most recent 7 years.

Select the appropriate geographic area and years. 

Group results by county. For Low Birth Weight births selects births 0-2499 grams. For all births (denominator) select all birth weights with the exception of those missing weight. 

CHR&R calculates data from raw NCHS databases and this matches CDC WONDER data. 

All raw NCHS birth data is saved on an external hard drive because it requiers a DUA. 

For the 2026 release, we use years 2018-2024 from NCHS. 


# import packages

```{r}
library(tidyverse)
library(haven)
library(here)
```


### standard county and state fips

The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.

```{r}
fips <-  bind_rows(
  read_sas(here("inputs/county_fips_with_ct_old.sas7bdat")) ,
  read_sas(here("inputs/state_fips.sas7bdat")))  %>% 
  arrange(statecode, countycode) %>% 
  glimpse()
```

# Load the NCHS births data 


```{r}
nat18 = readr::read_fwf("E:/CHRR/natality raw/rawdata/NATL2018US.AllCnty.txt", 
                         col_positions = readr::fwf_cols(bw = c(511), mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat19 = readr::read_fwf("E:/CHRR/natality raw/rawdata/nat2019us.AllCnty.txt", 
                         col_positions = readr::fwf_cols(bw = c(511), mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat20 = readr::read_fwf("E:/CHRR/natality raw/rawdata/NATL2020us.AllCnty.txt", 
                         col_positions = readr::fwf_cols(bw = c(511), mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat21 = readr::read_fwf("E:/CHRR/natality raw/rawdata/nat2021us.AllCnty.txt", 
                         col_positions = readr::fwf_cols(bw = c(511), mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat22 = readr::read_fwf("E:/CHRR/natality raw/rawdata/NATL2022US.AllCnty.txt", 
                         col_positions = readr::fwf_cols(bw = c(511), mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat23 = readr::read_fwf("E:/CHRR/natality raw/rawdata/nat2023us.AllCnty.txt", 
                         col_positions = readr::fwf_cols(bw = c(511), mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat24 = readr::read_fwf("E:/CHRR/natality raw/rawdata/nat2024us.AllCounties.txt", 
                         col_positions = readr::fwf_cols(bw = c(511), mracehisp = c(117), county_fips = c(91, 93), state = c(89,90), mage = c(79)))

nat = list(nat18, nat19, nat20, nat21, nat22, nat23, nat24)

#remove births with mother's residence American Samoa, Guam, Northern Marianas, Puerto Rico, Virgin Islands, not applicable, and not classified ;
nat_fixed = lapply(nat, function(x) x %>% filter(!(state %in% c("AS", "GU", "MP", "PR", "VI", "XX", "ZZ", "AB", "BC", "MB", "NB", "ON", "QC", "SK", "ZZ")) & !(county_fips %in% c("000", "999"))) %>% mutate(birth = 1))

# get county level estimates 
nat_fixed_sum = lapply(nat_fixed, function(x) x %>% group_by(state, county_fips, bw) %>% summarise(tbirth = sum(birth)))

#turn list into a single dataframe with numeric cols 
all_births = nat_fixed_sum %>% reduce(full_join, by = c("state", "county_fips", "bw")) 
all_births = data.frame(all_births)
all_births[4:10] <- sapply(all_births[4:10],as.numeric)


#calculate total births over seven year period
all_births$birthsums = rowSums(all_births[4:10], na.rm = TRUE)

#get state codes with abbrevs 
state = tigris::fips_codes %>% select(state, state_code) %>% distinct()

#add in state abbrevs 
all_births_state = merge(all_births, state, by = "state")

#verify that nothing important got dropped 
nojoin = dplyr::anti_join(all_births, all_births_state, by = "state")



#remove all bw=4 rows since bw=4 indicates that weight was not recorded 
all_births_fixed = all_births_state %>% filter(state != "VI") %>% filter(bw !=4 & !is.na(bw)) %>% select("state_code", "county_fips", "birthsums", "bw") 


#fix some fips that changed: 
# 02270, 46113, 51515

all_births_fixed2 = all_births_fixed %>% mutate(county_fips = case_when(
      state_code =="46" & county_fips =='113' ~'102',
      state_code =='02' & county_fips =='270' ~'158',
      state_code =='51' & county_fips =='515' ~'019',
      TRUE ~ county_fips
      ))


# calculate county level numerator and denominator for low birth weight calcs   
lbw_num_denom = all_births_fixed2 %>% group_by(state_code, county_fips) %>% 
  summarize(num = sum(birthsums[bw %in% c(1,2)], na.rm = TRUE),
            denom = sum(birthsums, na.rm = TRUE))

```



# calculate state and national numerators and denominators 

```{r}
# state 
statesums = all_births_fixed2  %>% group_by(state_code) %>% 
  summarize(num = sum(birthsums[bw %in% c(1, 2)], na.rm = TRUE),
            denom = sum(birthsums, na.rm = TRUE))
statesums$county_fips = "000"

#national
natsum = all_births_fixed2 %>%
  summarize(num = sum(birthsums[bw %in% c(1,2)], na.rm = TRUE),
            denom = sum(birthsums, na.rm = TRUE))
natsum$state_code = "00" 
natsum$county_fips = "000"

lbw_all = rbind(lbw_num_denom, statesums, natsum)

# apply suppression criteria 
lbw_all$num = ifelse(lbw_all$num < 10, NA, lbw_all$num)
lbw_all$denom = ifelse(lbw_all$num < 10, NA, lbw_all$denom)



```

# Calculate rates and confidence intervals 
The methods for calculating the error associated with measures of natality can be found here:

https://ftp.cdc.gov/pub/Health_Statistics/NCHS/Dataset_Documentation/DVS/natality/UserGuide2014.pdf

Since Low Birth Weight is measured as a percentage of all live births we are able to calculate the 95% error margins using the following equations:

95% high=county estimate+1.96*sqrt((1-county estimate)*(county estimate)/number of births)

95% low=county estimate-1.96*sqrt((1-county estimate)*(county estimate)/number of births)


```{r}

lbw_all$rate = lbw_all$num/ lbw_all$denom


#need standard error first 
#rse of a proportion = sqrt(p*(1-p)/n) / p 

lbw_all = lbw_all %>% group_by(state_code, county_fips) %>% 
  mutate(rse = sqrt(rate*(1-rate)/ denom)/rate)

lbw_all_ci = lbw_all %>% 
  mutate(lci = ifelse(!(num %in% 0:9), rate - 1.96*sqrt((1-rate)*rate/denom), NA), #suppress CIs when fewer than 10 lbw births 
         uci = ifelse(!(num %in% 0:9), rate + 1.96*sqrt((1-rate)*rate/denom), NA), 
         unreliable = ifelse(rse >0.2 | num <20, 1, 0))



# add back original fipscodes and rename cols to include measure number 

v037 = merge(fips, lbw_all_ci, by.x = c("statecode", "countycode"), by.y = c("state_code", "county_fips"), all.x = TRUE) %>% 
  rename(v037_numerator = num,
         v037_denominator = denom,
         v037_rawvalue = rate, 
         v037_cilow = lci,
         v037_cihigh = uci)


```


Now verify calculations 
```{r}
source(here("scripts/nchs_measure_helpers.R"))
compare("v037", v037, 2026)
```


#save to project 
```{r}
readr::write_csv(v037, file = here("measure_datasets/v037_R2026.csv"))

```




