---
title: "v167 school segregation"
author: "Hannah Olson-Williams"
output: html_document
---

# Description 
The extent to which students within different race and ethnicity groups are unevenly distributed across schools when compared with the racial and ethnic composition of the local population. The index ranges from 0 to 1 with lower values representing a school composition that approximates race and ethnicity distributions in the student populations within the county, and higher values representing more segregation.

# Connection to health 
The Brown v. Board of Education United States Supreme Court case declared racial segregation in public schools unconstitutional in 1954. Progress toward desegregation was slow and opposed by many individuals, organizations, and government officials.1 For example, in the late 1950s Native students of the Poarch Band of Creek Indians were permitted to ride buses to previously all-white schools in Alabama, but only if the Native students had skin deemed light enough.2 Desegregation of schools peaked in the late 1980s and early 1990s. In 1991, in the Board of Education of Oklahoma City v. Dowell decision, the Supreme Court ruled that desegregation efforts may be dissolved. As a result of such legal rulings and government policies and decisions that support segregated living and school enrollment, contact between racial groups in public schools has declined.3 Approximately 37% of Black students in the U.S. enrolled in majority white schools in 1988, but in 2018, the percentage declined to 19%.3 For Hispanic students, school segregation has steadily increased with population shifts and Hispanic students now face higher levels of segregation nationally than Black students.3

The Brown decision deemed the doctrine of “separate but equal” unconstitutional and acknowledged that equality in segregation is not possible. This ruling is supported by recent statistics showing that Black and Hispanic students are most likely to experience intense segregation in school and also most likely to be concentrated in under-resourced, high-poverty schools with less experienced and credentialed teachers and higher teacher turnover.3-5 In rural areas white and Native students can face high levels of segregation and similar barriers to success, including increased poverty.6

School segregation strongly reflects residential segregation; however, the policies that drive school and residential segregation are under the jurisdictions of separate decision-making bodies, creating multiple avenues for progress. School segregation is more highly associated with achievement gaps than residential segregation, which suggests that school segregation may affect additional pathways to student achievement beyond those explained by residential segregation (e.g., teacher retention and credentials, school resources, breadth of curriculum, and advanced course offerings).4

Schools where the majority of students have minoritized racial or ethnic identities can be culturally affirming by centering ways of being for students who may otherwise find themselves on the margins; however, demonstrated relationships between economic segregation, racial segregation, and fewer academic opportunities mean that students in these settings may still experience inequitable outcomes.7 Studies show that racial/ethnic segregation in schools is negatively associated with achievement, college success, long-term employment, and income for students of color. School segregation leads to racial gaps in the identification of students who would benefit from special education, perpetuating racial gaps into adulthood.8

In addition to negatively impacting opportunities for students in schools experiencing disinvestment, school segregation can hinder the development of diverse perspectives. In the U.S., social networks continue to be homogeneous - for example, the majority of white adults in the U.S. have entirely white social networks.9 While not without challenges, studies have found that students in diverse, well-resourced schools have more opportunities to gain civic benefits, including skills in working collaboratively across lines of difference, development of equitable and inclusive habits, and experience in aggregating wisdom.10

# School segregation is an index 
School Segregation measures how evenly representation of racial and ethnic groups in the student population is spread across schools using Theil's Index, a segregation index. The index ranges from 0 to 1 with lower values representing a school composition that approximates race and ethnicity distributions in the student populations within the county, and higher values representing more segregation.

# Some data are suppressed 
An estimate is only calculated when at least two of the race-ethnicity categories have 25 or more students in at least one school within a given county.

# Caution should be used when comparing these estimates across years
There are inconsistencies between states in how students in schools run by Bureau of Indian Education (BIE) are counted. In states that provide support to federal BIE schools, students at these schools may be double counted in National Center for Education Statistics (NCES) data. 

# Measure limitations 

NCES data is limited in that it does not include private school students. According to the National Center for Education Statistics, 11% of students attended private schools in 2019. The opportunity for a given student to attend a school across districts is not independent of race. Private schools in the U.S. disproportionately serve white and higher-income families. 

School segregation can occur at a variety of levels (e.g., school, district, county, region), by measuring at the county-level we are inevitably underreporting the amount of school segregation taking place in certain regions of the country. For instance, if school segregation is largely occurring between counties, and not within counties, this measure will vastly underestimate the true burden.

Multiple factors lead to the negative outcomes associated with school segregation. This measure cannot capture the interplay between school resources, economic segregation, racial segregation, and academic achievement. 


# Website to downloaded data 
https://nces.ed.gov/ccd/elsi/tableGenerator.aspx


# Directions
Select a Table Row: "Public School"

Select Years: "2023-2024"

Select Table Columns: 

Information/Basic Information
State Name [Public School] Latest available year
School Name [Public School] 2023-24
County Name [Public School] 2023-24
County Number [Public School] 2023-24

Characteristics/School-District Classification Information
School Type [Public School] 2023-24
Start of Year Status [Public School] 2023-24

Enrollments/Total Enrollment
Total Students All Grades (Excludes AE) [Public School] 2023-24

Enrollments/Enrollment by Race/Ethnicity
American Indian/Alaska Native Students [Public School] 2023-24
Asian or Asian/Pacific Islander Students [Public School] 2023-24
Hispanic Students [Public School] 2023-24
Black or African American Students [Public School] 2023-24
White Students [Public School] 2023-24
Nat. Hawaiian or Other Pacific Isl. Students [Public School] 2023-24
Two or More Races Students [Public School] 2023-24

Select Filters (Refinements): Can leave without altering.

Click "Create Table"
Click CSV 
Click Download Zip File 

# Calculating this measure 
### State value
population-based (internal calculation)

### National value 
median of states 

### state/national value 
national value calculated as median of state values 

1. The following school types are excluded:
if find(school_name, 'JUVENILE') then delete;
if find(school_name, 'JAIL') then delete;
if find(school_name, 'PRISON') then delete;
if find(school_name, 'CORRECTIONAL') then delete;
if find(school_name, 'VIRTUAL') then delete;
if find(school_name, 'CYBER') then delete;
where School_Type__Public_School__2023 ne '4-Alternative Education School';
where Start_of_Year_Status__Public_Sch ne '6-Inactive';
if Start_of_Year_Status__Public_Sch = '2-Closed' then delete;
also, if a school's Total Student population = 0 then delete

2. Create proportions for each race/ethnicity category:
amer_ind_prop = AIAN_students/Total_Students
asian_pi_haw_prop = (Asian_PI_students+Hawaiian_students)/Total_Students
hispanic_prop = Hispanic_students/Total_Students
black_prop = Black_students/Total_Students
white_prop = White_students/Total_Students
two_prop = Two_or_more_students/Total_Students
*set student populations by race/ethnicity to 0 that are set to missing by NCES;

3. Calculate a theil value for each school:
theil_school = -(amer_ind_prop * log(ifn(amer_ind_prop=0, 1, amer_ind_prop)) + asian_pi_haw_prop * log(ifn(asian_pi_haw_prop=0, 1, asian_pi_haw_prop)) + hispanic_prop * log(ifn(hispanic_prop=0, 1, hispanic_prop)) + black_prop * log(ifn(black_prop=0, 1, black_prop)) + white_prop * log(ifn(white_prop=0, 1, white_prop)) + two_prop * log(ifn(two_prop=0, 1, two_prop)))

(function ifn() is used to avoid log0)

4. Calculate a theil value for each county:
theil_county = aggregate school values for each race/ethnicity category and complete the same calculation as above for theil_school

5. Calculate a school_weight, theil_school_weight, and a theil_county_weight:
school_weight = Total_students_in_school/Total_students_in_county
theil_school_weight = theil_school*school_weight
theil_county_weight = aggregating the theil_school_weight for each county

6. Use the below formula to create the final theil value for each county:
theil_county_final = abs(theil_county-theil_county_weight)/theil_county
Then suppress counties where only one race/ethnicty category has 25 or more students or where there is only one school

7. Follow the same process as above for calculating state values. It might be easiest to start with the dataset that is used to aggregate student populations by race/ethnicity group from school-level to county-level. Then follow steps 2-6 above.



# import packages

```{r}
library(tidyverse)
library(haven)
library(here)
```


### standard county and state fips

The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.

```{r}
fips <-  bind_rows(
  read_sas(here("inputs/county_fips_with_ct_old.sas7bdat")) ,
  read_sas(here("inputs/state_fips.sas7bdat")))  %>% 
  arrange(statecode, countycode) %>% 
  glimpse()
```

# Load raw data 

```{r}
elsi_raw = read.csv(here("raw_data/ELSI/ELSI_csv_export.csv"), skip = 6)
```

# format, filter, calculate 

```{r}
elsi_formatted = elsi_raw %>%
  filter(!grepl("JUVENILE|JAIL|PRISON|CORRECTIONAL|VIRTUAL|CYBER", School.Name, ignore.case = TRUE)) %>% 
  filter(School.Type..Public.School..2023.24 != "4-Alternative Education School") %>% 
  filter(!(Start.of.Year.Status..Public.School..2023.24 %in% c("6-Inactive", "2-Closed"))) %>% 
  filter(Total.Students.All.Grades..Excludes.AE...Public.School..2023.24 != 0) %>% 
  filter(!(Total.Students.All.Grades..Excludes.AE...Public.School..2023.24 %in% c("–","†", "‡"))) %>% #need to add this to admin 
  filter(Start.of.Year.Status..Public.School..2023.24 != "7-Future") %>% #need to add this to admin 
  filter(!is.na(County.Number..Public.School..2023.24)) %>% #gotta add to admin 
  mutate(fipscode = str_pad(County.Number..Public.School..2023.24, width = 5, side = "left", pad = "0"), 
         statecode = substr(fipscode, 1,2), 
         countycode = substr(fipscode, 3,5),
         
         # Replace special characters such as '†' and '‡' and '-' with NA
         n_school = str_replace_all(Total.Students.All.Grades..Excludes.AE...Public.School..2023.24, "[†‡-]", NA_character_),
         n_amer_ind = str_replace_all(American.Indian.Alaska.Native.Students..Public.School..2023.24, "[†‡-]", NA_character_),
         n_haw = str_replace_all(Nat..Hawaiian.or.Other.Pacific.Isl..Students..Public.School..2023.24, "[†‡-]", NA_character_),
         n_asian_pi = str_replace_all(Asian.or.Asian.Pacific.Islander.Students..Public.School..2023.24, "[†‡-]", NA_character_),
         n_hispanic = str_replace_all(Hispanic.Students..Public.School..2023.24, "[†‡-]", NA_character_),
         n_black = str_replace_all(Black.or.African.American.Students..Public.School..2023.24, "[†‡-]", NA_character_),
         n_white = str_replace_all(White.Students..Public.School..2023.24, "[†‡-]", NA_character_),
         n_two = str_replace_all(Two.or.More.Races.Students..Public.School..2023.24, "[†‡-]", NA_character_),
         
         
         
         # Convert to numeric and replace NA with 0
         n_school = coalesce(as.numeric(n_school), 0), 
         n_amer_ind = coalesce(as.numeric(n_amer_ind), 0),
         n_asian_pi_haw = coalesce(as.numeric(n_asian_pi),0) + coalesce(as.numeric(n_haw),0),
         n_hispanic = coalesce(as.numeric(n_hispanic), 0),
         n_black = coalesce(as.numeric(n_black), 0),
         n_white = coalesce(as.numeric(n_white), 0),
         n_two = coalesce(as.numeric(n_two), 0),
         
         # calculate proportions
         amer_ind_prop = n_amer_ind/n_school,  
         asian_pi_haw_prop = n_asian_pi_haw/n_school, 
         hispanic_prop = n_hispanic/n_school,  
         black_prop = n_black/n_school, 
         white_prop = n_white/n_school, 
         two_prop = n_two/n_school,
         
         # calculate school level theils 
         theil_school = -(amer_ind_prop * log(ifelse(amer_ind_prop==0, 1, amer_ind_prop)) + 
                            asian_pi_haw_prop * log(ifelse(asian_pi_haw_prop==0, 1, asian_pi_haw_prop)) + 
                            hispanic_prop * log(ifelse(hispanic_prop==0, 1, hispanic_prop)) + 
                            black_prop * log(ifelse(black_prop==0, 1, black_prop)) + 
                            white_prop * log(ifelse(white_prop==0, 1, white_prop)) + 
                            two_prop * log(ifelse(two_prop==0, 1, two_prop)))
         ) 
```

# use school-level theils to calculate county-level theils 
```{r}
theil_county = elsi_formatted %>% group_by(statecode, countycode) %>% 
  mutate(
    
    nc_totalstudents = sum(n_school, na.rm = TRUE),
    nc_amer_ind = coalesce(sum(n_amer_ind, na.rm = TRUE), 0),  
    nc_asian_pi_haw = coalesce(sum(n_asian_pi_haw, na.rm = TRUE), 0),  
    nc_hispanic = coalesce(sum(n_hispanic, na.rm = TRUE), 0),   
    nc_black = coalesce(sum(n_black, na.rm = TRUE), 0),  
    nc_white = coalesce(sum(n_white, na.rm = TRUE), 0),  
    nc_two = coalesce(sum(n_two, na.rm = TRUE), 0), 
    
    amer_ind_propc = nc_amer_ind/nc_totalstudents,  
    asian_pi_haw_propc = nc_asian_pi_haw/nc_totalstudents, 
    hispanic_propc = nc_hispanic/nc_totalstudents,  
    black_propc = nc_black/nc_totalstudents, 
    white_propc = nc_white/nc_totalstudents, 
    two_propc = nc_two/nc_totalstudents, 
    
    theil_county =  -(amer_ind_propc * log(ifelse(amer_ind_propc==0, 1, amer_ind_propc)) + 
                        asian_pi_haw_propc * log(ifelse(asian_pi_haw_propc==0, 1, asian_pi_haw_propc)) + 
                        hispanic_propc * log(ifelse(hispanic_propc==0, 1, hispanic_propc)) + 
                        black_propc * log(ifelse(black_propc==0, 1, black_propc)) + 
                        white_propc * log(ifelse(white_propc==0, 1, white_propc)) + 
                        two_propc * log(ifelse(two_propc==0, 1, two_propc)))
      )
```


# calculate weights 
```{r}
theil_weights = theil_county %>% mutate(
  school_weight = n_school / nc_totalstudents,
  theil_school_weight = theil_school*school_weight 
) 
```


# calculate raw values 
```{r}
theil_final = theil_weights %>% 
  group_by(statecode, countycode) %>% 
  mutate(
    theil_county_weight = sum(theil_school_weight, na.rm = TRUE),
    theil_county_final = ifelse(
      rowSums(cbind(nc_amer_ind >= 25, nc_asian_pi_haw >= 25, nc_hispanic >= 25, 
                                               nc_black >= 25, nc_white >= 25, nc_two >= 25)) >1 & 
        n()>1,
                                abs(theil_county - theil_county_weight)/theil_county, 
      NA)
    )

# remove school-level estimates to leave only county-level estimates 
v167_school_segregation = theil_final %>% group_by(statecode, countycode) %>% select(theil_county_final) %>% distinct()

```

# add in standardized fipscodes and rename  
```{r}
v167_school_segregation_fips = merge(fips, v167_school_segregation, by = c("statecode", "countycode")) %>% 
  rename(v167_rawvalue = theil_county_final)
```




#save to project 
```{r}
readr::write_csv(v167_school_segregation_fips, file = here("measure_datasets/v167_R2025.csv"))

```




